Pragmatic interactions lead to efficient language structure and use

### 2019-01-28 paper/paper.Rmd
updated paper and supplementary materials after M Hahn comments.

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game (similar to those described by Wittgenstein). We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1 for an example reference game). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$. Note that the prior over meanings are analogous to the *need probabilities* assumed in previous work Regier, Kemp & Kay [-@RegierKempKay2015a]. We consider a set of contexts $C$ with an associated prior $P(C)$. Contexts describe different orderings of the underlying meaning prior, formalized as different conditional distributions over meanings e.g. $p(M|C=c_i)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meaning pairs.\par

### Languages
A language $\ell$ defines the set of semantic mappings between utterance and meanings. For example, in a world with three utterances $U = \{u_1, u_2, u_3\} \text{ and three meanings }M = \{m_1, m_2, m_3\}$ the boolean matrix

\begin{center}
\begin{tabular}{ c | c c c } 
& $m_1$ & $m_2$ & $m_3$ \\
\hline
$u_1$ & 1 & 1 & 0 \\
$u_2$ & 0 & 1 & 0 \\
$u_3$ & 0 & 0 & 1 \\
\end{tabular}
\end{center}

describes the literal semantics of $\ell$.

@@

\begin{equation}
  \text{Efficiency}(S, L, \ell) = \mathbb{E}_{e \sim P(E)}[\text{speaker effort}] + \mathbb{E}_{e \sim P(E)}[\text{listener effort}]
\end{equation}

We assume that speaker effort is the negative log probability of an utterance $-log_2(p(U))$, listener effort is the negative log probability of the listener's conditional $-log_2(L(M|U))$, and that events $e = <u, m>$ are sampled according the to following generative model -- some object occurs in the world with probability $p(M=m)$, the speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m)$ (i.e. $e \sim p(m)S(u|m)$). Using these basic ingredients we arrive at the following objective:

\begin{equation}
\begin{split}
  = \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(P_{listener}(u, m; \ell))]
\end{split}
\end{equation}
This is the simply the cross-entropy between the speaker and listener joint distributions.
\begin{equation}
\begin{split}
  = \mathbb{E}_{P_{speaker}}[-log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener})
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive. Cross-entropy gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, this is the difference between the joint distribution assumed by the speaker $P_{speaker}$ and listener $P_{listener}$. An "efficient" language $\ell$ used by a set of a pair of speaker-listeners will have properties which minimize this objective. For a complete derivation and its componetns we refer the reader to Section 2.1 of our supplementary materials. \par

@@

We conduct $N=2000$ simulations. For each simulation we enumerate the set of *valid* languages in which $|U|=|M|=4$. Note that a language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment) and each form maps to at least one meaning (every column has some non-zero assignment). For a given simulation the goal is to find the language $\ell^*$ which minimizes our cross-entropy objective.\par

>>>>

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game (similar to those described by Wittgenstein). We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1 for an example reference game). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}. We consider a set of contexts $C$ with an associated prior $P(C)$. Contexts describe different orderings of the underlying meaning prior, formalized as different conditional distributions over meanings e.g. $p(M|C=c_i)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meaning pairs.\par

### Languages
A language $\ell$ defines the set of semantic mappings between utterance and meanings. For example, figure 1 contains four utterances $U = \{\text{blue}, \text{green}, \text{square}, \text{circle}\} \text{ and three meanings }M = \{\text{green-square}, \text{blue-square}, \text{green-circle}\}$. The boolean matrix describes the literal semantics of the language $\ell$. Note that will consider language to "contain ambiguous items" if there is at least one utterance $u \in U$ which can apply to multiple meanings (i.e. $|[[u_i]]| > 1$). For example, in Figure 1 both the words "blue" and "green" are ambigous so we would say that $\ell$ contains ambiguity.

@@

\begin{equation}
\begin{split}
  \text{Efficiency}(S, L, \ell) = \mathbb{E}_{e \sim P(E)}[\text{speaker effort}] \\+ \mathbb{E}_{e \sim P(E)}[\text{listener effort}]
\end{split}
\end{equation}

We assume that speaker effort is the negative log probability of an utterance (surprisal) -- intuitively, the number of bits needed to encode the utterance $u$. This particular formalization of speaker-cost is general enough to accommodate a range of instantiations that might in theory be related to production difficult via articulation effort, cognitive effort related to lexical access, etc [@BennettGoodman2015a].\par

$$\text{speaker effort} = -log_2(p(u))$$

We assume listener effort is the negative log probability of the listener's conditional $-log_2(L(M|U))$. This operationalization of listener effort is intuitively related to existing work in sentence processing in which word difficulty during comprehension is proportional to surprisal [@Hale2001a; @Levy2008a].

$$\text{listener effort} = -log_2(L(m|u; \ell))$$

Finally, we assume that events $e = <u, m>$ are sampled according the to following generative model -- some object occurs in the world with probability $p(M=m)$, the speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m)$ (i.e. $e \sim p(m)S(u|m)$). Using these basic ingredients we arrive at the following objective:

\begin{equation}
\begin{split}
  = \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(P_{listener}(u, m; \ell))]
\end{split}
\end{equation}
This is the simply the cross-entropy between the speaker and listener joint distributions.
\begin{equation}
\begin{split}
  = \mathbb{E}_{P_{speaker}}[-log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener}; \ell)
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive. Cross-entropy gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, this is the difference between the joint distribution assumed by the speaker $P_{speaker}$ and listener $P_{listener}$. An "efficient" language $\ell$ used by a set of a pair of speaker-listeners will have properties which minimize this objective. For a complete derivation and its components we refer the reader to Section 2.1 of our supplementary materials.\par

@@

We conduct $N=2000$ simulations. For each simulation we enumerate the set of *valid* languages in which $|U|=|M|=4$. Note that a language $\ell \in L$ is "valid" so long as each possible meaning in $m \in M$ can be referred to by at least one form $u \in U$ (every column of $\ell$ has some non-zero assignment) and each form maps to at least one meaning (every column has some non-zero assignment). For a given simulation the goal is to find the language $\ell^*$ which minimizes our cross-entropy objective and then check to see if that language contains ambiguity.\par
